Phylogenetic comparative methods in the lme4-verse

### 2019-04-09 example.rmd
pushing branch

```{r pkgs,message=FALSE, warning=FALSE}
library(ape)
library(lme4)
library(Matrix)
library(MCMCglmm)
library(MASS) 
library(pez)
library(glmmTMB)
library(dplyr)
library(coda)
library(lattice)
library(broom)
library(dotwhisker)
```

@@

An alternative approach is to model the phylogenetic correlation as a *Gaussian process*. In particular, suppose that the evolutionary process is a Brownian motion (an almost certainly incorrect/oversimplified model of evolution, but one that many phylogenetic methods are built on). In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. If we set up the $Z$ matrix appropriately, we can model everything with a sequence of *independent* errors, rather than having to do fancy things to impose a correlation structure on the random effects.

@@

```{r setup}
library(ape)
set.seed(101)
r <- makeNodeLabel(rtree(4))
plot(r,show.node.label=TRUE)
```

@@

Here's a (probably not very efficent) way to construct the Z matrix.
(There must be a way to not walk the tree multiple times from every tip ...

```{r phylo.to.Z}
phylo.to.Z <- function(r,stand=FALSE){
  ntip <- length(r$tip.label)
  Zid <- Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
  nodes <- (ntip+1):max(r$edge)
  root <- nodes[!(nodes %in% r$edge[,2])]
  for (i in 1:ntip){
    cn <- i  ## current node
    while (cn != root){
      ce <- which(r$edge[,2]==cn)   ## find current edge
      Zid[i,ce] <- 1   ## set Zid to 1
      cn <- r$edge[ce,1]            ## find previous node
    }
  }
  V <- vcv(r)
  # V <- V/max(V)
  sig <- exp(as.numeric(determinant(V)["modulus"])/ntip)
  # sig <- det(V)^(1/ntip)
  Z <- t(sqrt(r$edge.length) * t(Zid))
  if(stand){Z <- t(sqrt(r$edge.length/sig) * t(Zid))}
  rownames(Z) <- r$tip.label
  colnames(Z) <- 1:length(r$edge.length)
  return(Z)                                  
}
phylo.to.Z(r)
```

@@

```{r phylocode}
source("lme4_phylo_setup.R")
```

>>>>

@@

An alternative approach is to model the phylogenetic correlation as a *Gaussian process* (GP). In particular, suppose that the evolutionary process is a Brownian motion (an almost certainly incorrect/oversimplified model of evolution, but one that many phylogenetic methods are built on). In that case, the phylogenetic variability of a particular observation can be written as the sum of the evolutionary changes that occurred on all of the branches in the phylogeny in its past. If we set up the $Z$ matrix appropriately, we can model everything with a sequence of *independent* errors, rather than having to do fancy things to impose a correlation structure on the random effects.

@@

```{r setup}
library(ape)
library(Matrix)
set.seed(101)
r <- makeNodeLabel(rtree(4))
plot(r,show.node.label=TRUE)
```

@@

Here's a (probably not very efficent) way to construct the Z matrix.
(There must be a way to not walk the tree multiple times from every tip ...)

```{r phylo.to.Z}
phylo.to.Z <- function(r,stand=FALSE){
  ntip <- length(r$tip.label)
  Zid <- Matrix(0.0,ncol=length(r$edge.length),nrow=ntip)
  nodes <- (ntip+1):max(r$edge)
  root <- nodes[!(nodes %in% r$edge[,2])]
  for (i in 1:ntip){
    cn <- i  ## current node
    while (cn != root){
      ce <- which(r$edge[,2]==cn)   ## find current edge
      Zid[i,ce] <- 1   ## set Zid to 1
      cn <- r$edge[ce,1]            ## find previous node
    }
  }
  V <- vcv(r)
  sig <- exp(as.numeric(determinant(V)["modulus"])/ntip)
  Z <- t(sqrt(r$edge.length) * t(Zid))
  if(stand){Z <- t(sqrt(r$edge.length/sig) * t(Zid))}
  rownames(Z) <- r$tip.label
  colnames(Z) <- 1:length(r$edge.length)
  return(Z)                                  
}
phylo.to.Z(r)
```

@@

```{r pkgs,message=FALSE, warning=FALSE}
library(ape)
library(lme4)
library(MCMCglmm)
library(MASS) 
library(pez)
library(glmmTMB)
library(dplyr)
library(coda)
library(lattice)
library(broom)
library(dotwhisker)
```

The phylo--to--Z function is already in the source code.
```{r phylocode}
source("lme4_phylo_setup.R")
```
